---
title: "Lab 2 & 3 Key"
---

Load necessary packages.
```{r packages, message = FALSE}
needs(tidyverse, rio, here, knitr, kableExtra, misty, janitor, rstatix, psych) 
```

Import data.
```{r}
df_2 <- misty::read.sav(here("assignments_data/Lab3_Vocab.sav"), 
                      use.value.labels = TRUE) %>% 
              clean_names() # cleaning up variable names
```

Calculate descriptive statistics.
```{r}
describe(df_2) %>% 
  select(n, mean, sd, skew, kurtosis, se)

# Histogram
ggplot(data = df_2, aes(x = vocab)) +
  geom_histogram(bins = 20)

# Boxplot
ggplot(data = df_2, aes(x = instruct, y = vocab)) +
  geom_boxplot()

# Homogeneity of variance
needs(car)
car::leveneTest(vocab ~ instruct, data = df_2, center = "mean")
```

Calculate F-ratio.
```{r}
needs(afex)
aov_car(vocab ~ instruct + Error(idnum), data = df_2)
```

Post-Hoc Comparison. Use GH.
```{r}
#post-hoc
rstatix::games_howell_test(vocab ~ instruct, data = df_2,
                           conf.level = 0.95,
                           detailed = FALSE) #outputs actual estimates with GH correction
# p.adj is GH adjusted p value
```

Post-hoc contrasts.

Helmert is best for this case. Physical vs others (1, -.5, -.5); physical vs history (0, 1, -1) = 0 + -.5 + .5 = 0 --> Orthogonal! 

```{r}
#check the order of levels
levels(df_2$instruct)

# change levels to a different order using the following code (but we don't actually need to here)
# df_2$instruct <- factor(df_2$instruct, levels = c("social science", "history", "physical science"))
```

Step 1. Set contrasts (Helmert)
```{r}
contrast1 <- c(1, -.5, -.5) #physical science vs others
contrast2 <- c(0, 1, -1) #social science vs history
```

Step 2. Bind vectors to temporary matrix, where constants are equal to 1/(length of vectors)
```{r}
mat.temp <- rbind(constant = 1/3, contrast1, contrast2) #3 is for number of cells in the design (same as levels because only one IV)
mat.temp
```

Step 3. Take the inverse of the matrix.
```{r}
mat <- solve(mat.temp)
mat

#drop first column of constants
mat <- mat[, -1]
mat
```

Step 4. Run model with lm() and set contrasts: link "instruct" variable with contrast matrix
```{r}
m_contrasts <- lm(vocab ~ instruct, data = df_2, contrasts = list(instruct = mat))

summary(m_contrasts)
```
Level 1 vs later is 9.75 (instruct 1): first half of the table from the slides = instruct1 output.

If we are only using these two (2) contrasts, use alpha of 0.05/2 = 0.025 as the new alpha. 

But, if you count the 3 pairwise comparisons already done in lab 2 a = 0.05/5 = 0.01. Use this since F test was significant.

So, we end up with a nonsignificant result: we fail to reject our null hypothesis. We cannot say the physical science condition is significantly different from the humanities conditions.

```{r}
anova_type3_ges <- anova_test(m_contrasts, 
                 type="III",
                 effect.size = "pes",
                 detailed = TRUE)

anova_type3_ges
```

Calculate Omega:
* dfn = 2
* dfd = 33
* Fvalue = 4.284
* n = 36
* a = .05

Now, just plug them in.
```{r}
needs(MOTE)
omega.F(dfm = 2, dfe = 33,
      Fvalue = 4.284, n = 36, a = .05)
```