---
title: 'EDUC 640: Quick Reference Guide'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
---

I will use the data from the `palmerpenguins` package to illustrate examples on here.
```{r}
library(palmerpenguins)
```


## General useful R functions

* use `?` or `??` before a function name to learn more about a function (also applies for a package)

* use `head()` to examine the first five rows of data
<input type=button class=hideshow></input>
```{r}
head(penguins)
```

* use `str()` to look at the structure of the dataframe (and see if variables are in the right format)
<input type=button class=hideshow></input>
```{r}
str(penguins)
```

* use `factor()` or `as.factor()` to convert (usually) a categorical variable into countable categories (ex. group 1, group 2, group 3 instead of a "string" type). Can also be used numbers that should be categories instead of a continuous integer (ex. if you are comparing years 2016 vs. 2017 or student 3 with student 7)
<input type=button class=hideshow></input>
```{r}
#species is already a factor, so this code is redundant
```

* can reorder levels using `relevel()`

```{r}

```


* use `clean_names()` to turn variable names all lowercase and into one word



## Functions useful for checking assumptions

* In the `psych` package...
  - use `describe()` for general descriptive statistics
  - use `describeBy()` when you have multiple independent variables and you want to "describe by" group

* In the `car` package...
  - use `leveneTest()` to test homogeneity of variance. Within brackets...
    - input the same formula as for the ANOVA you want to run
    - specify `center = "mean"` (function's default is median)

## All about `ggplot()`

* histogram examples:

* boxplot examples:

## Functions for data wrangling

https://stackoverflow.com/questions/42630248/collapsing-hiding-figures-in-r-markdown


<script>
$( "input.hideshow" ).each( function ( index, button ) {
  button.value = 'Hide Output';
  $( button ).click( function () {
    var target = this.nextSibling ? this : this.parentNode;
    target = target.nextSibling.nextSibling.nextSibling.nextSibling.nextSibling;
    if ( target.style.display == 'block' || target.style.display == '' ) {
      target.style.display = 'none';
      this.value = 'Show Output';
    } else {
      target.style.display = 'block';
      this.value = 'Hide Output';
    }
  } );
} );
</script>
