---
title: "Lab 2 & 3"
---

# Introduction

This tutorial will extend the analysis using one-way ANOVA from Lab 1, specifically in illustrating pairwise comparisons, generating contrasts, and calculating effect size.

We will use data from your textbook (see ['Support Materials' on the textbook webpage](https://www.routledge.com/Design-and-Analysis-in-Educational-Research-ANOVA-Designs-in-SPSS/Strunk-Mwavita/p/book/9781138361164)) (Strunk & Mwavita, 2020). For your Lab 2 assignment, you will use the data provided on Canvas.

First, load necessary packages:
```{r packages, message = FALSE}
needs(tidyverse, rio, here, knitr, kableExtra, misty, janitor, rstatix, psych)
```

Next, we will read in the data. Remember to load the `misty` package for `.sav` files and `use.value.labels = TRUE` to get the labels as the cell values.
``` {r}
needs(misty)
df <- misty::read.sav(here("kamden-mwavita_data/one-way ANOVA - Fischer et al.sav"), use.value.labels = TRUE) %>% 
  clean_names()

head(df)
```

# Lab 2

Now we're ready to move onto examining data and running our analyses. In this tutorial, `school_funding` will be our DV and `group` will be our independent variable (IV), like from Lab 1.

See the Lab 1 tutorial for how to check descriptive statistics and test the three ANOVA assumptions:

**1. Homogeneity of variance**: Variances between the groups should be the same.  
**2. Independence of observations**: Observations should be independent of one another.  
**3. Normality**: The distribution of the dependent variable should be normal.

## Computing the F Statistic

In the last lab, we learned how to compute the F statistic by hand. Luckily, we don't have to keep doing that! In the `afex` package, R has the `aov_car()` function that can compute the F-ratio for you.

This function works specifically for data in the long format. Type `?aov` to learn more about the function and how it can be used. In this case, plug in the formula (DV ~ IV) and add in the Error term.
```{r}
needs(afex)
aov_car(school_funding ~ group + Error(administrator_support), data = df)
```

The omnibus F is statistically significant (F(2, 635)=4.50, p=.011). Do these values match those you calculated in Lab 1? (Hint: ... yes)

## Conduct Follow-Up Pairwise Comparisons

Depending on your needs, you can conduct a number of pairwise comparisons in R. The most common are in the `rstatix` package: Tukey, Games-Howell, and Bonferroni. Each are different corrections for the multiple comparisons family-wise error rate and results in a p-adjusted value (`p.adj`) to examine.

Each of these corrections can be calculated by inputing the data and the formula; you should select which comparison works for you based on your data. 

Below is the code to compare each of these, but you will normally only choose one after this lab.
```{r}

#Tukey (usually more powerful when testing large numbers of means; generally preferred)
tukey_hsd(df, formula = school_funding ~ group)

#Bonferroni (usually more powerful when number of comparisons is small)
pairwise_t_test(df, formula = school_funding ~ group,
                p.adjust.method = "bonferroni")

#Games-Howell (used if assumption of variance homogeneity is violated)
games_howell_test(df, formula = school_funding ~ group, 
                  conf.level = 0.95,
                  detailed = FALSE)

```


What differences do you see between between the test outputs? Which would you use for your lab?

# Lab 3

## Generating contrasts

In this case, since you have already run your analysis, contrasts are post-hoc, as well. At this point, you have a few contrast options:

* Deviation - Is each group mean significantly different from the grand mean?
* Simple - Is each group mean different from the first group or last group?
* Difference - Is each group mean significantly different from the previous group mean? 2 vs 1, 3 vs 2, etc.
* Repeated - (opposite of Difference) 1 vs 2, 2 vs 3 
* Helmert - Is each group mean significantly different from all later group means? i.e. 1 vs (2, 3, 4)
* Polynomial - Is there a linear, quadratic, cubric trend?

For this lab, we will demonstrate a Helmert contrast: compare one condition to other groups/conditions combined. For example, does the better-than-expected group significantly different from both the lower-than-expected and as-expected groups?

First, check the order of levels so they are in the order that you want.

```{r}
levels(df$group)

# change levels to a different order using the following code (but we don't actually need to here)
# df$group <- factor(df$group, levels = c("Lower-than-expected", "As-expected", "Better-than-expected"))
```

Since the order in the dataset looks good, we will now hand-code the contrasts.

**Step 1.** Set contrasts (Helmert)
```{r}
contrast0 <- c(1, -.5, -.5) #better-than-expected vs others
```

**Step 2.** Bind vectors to temporary matrix, where constants are equal to 1/(length of vectors)
```{r}
mat.temp <- rbind(constant = 1/3, contrast0) #denominator 3 is for number of cells in the design (same as levels because only one IV)
mat.temp
```

**Step 3.** Take the inverse of the matrix.
```{r}
mat <- MASS::ginv(mat.temp)
mat

#drop first column of constants
mat <- mat[, -1]
mat
```

**Step 4.** Run model with lm() and set contrasts: link "group" variable with contrast matrix
```{r}
m_contrasts <- lm(school_funding ~ group, data = df, contrasts = list(group = mat))
#Rename (name of intercept, contrast, and everything else unnamed) to make output easier to interpret
attr(m_contrasts$coefficients, "names") <- c("Intercept", "Better-Expected vs. Others", "")
summary(m_contrasts)
```
*Remember your alphas:*
If we are only using these two (2) contrasts, use alpha of 0.05/2 = 0.025 as the new alpha. 

If you count the 3 pairwise comparisons already done in lab 2 a = 0.05/5 = 0.01. 

## Generating effect sizes

While `aov_car` generates effect sizes in the code, another way to run the ANOVA is with the code below. While we will be using ANOVA type 3 in this course, this code will allow you to change that if needed. It also lets you calculate different effect sizes, i.e. partial eta squared (pes) instead of general eta squared (ges) that automatically is calculated with `aov_car` above.

See below for calculating the effect size on the contrasts.

```{r}
anova_type3_ges <- anova_test(m_contrasts, 
                 type="III",
                 effect.size = "pes",
                 detailed = TRUE)

anova_type3_ges
```

To calculate omega squared, we can use MOTE package’s `omega.F()` function.

Omega squared or partial omega squared is calculated by subtracting one from the F-statistic and multiplying it by degrees of freedom of the model. This is divided by the same value after adding the number of valid responses. This value will be omega squared for one-way ANOVA designs, and will be partial omega squared for multi-way ANOVA designs (i.e. with more than one IV).

$omega^2 = (dfm * (Fvalue-1)) / ((dfm * (Fvalue-1)) + n)$

In this case, we need the following values (see ANOVA output for the corresponding values):

* dfn = 2
* dfd = 635
* Fvalue = 4.499
* n = 638
* a = .05

Now, just plug them in.
```{r}
needs(MOTE)
omega.F(dfm = 2, dfe = 635,
      Fvalue = 4.499, n = 638, a = .05)
```

