---
title: 'Lab 5: Two-way Factorial ANOVA'
---
## Introduction

This tutorial will learn how to run two-way factorial ANOVA. This lab will be a little shorter than those before -- woo! We will calculate descriptive statistics, check ANOVA assumptions, calculate ANOVA and effect sizes, and code contrasts according to our research questions.

We will use data from your textbook (see ['Support Materials' on the textbook webpage](https://www.routledge.com/Design-and-Analysis-in-Educational-Research-ANOVA-Designs-in-SPSS/Strunk-Mwavita/p/book/9781138361164)) (Bossaert et al., 2011). This data focuses on the social dimension of inclusion in education. The columns include the following information:

* `country`: name of country
* `soc_part`: social participation (score)
* `sen`: students with special educational needs categorization

This data is similar to the one you will use for your Lab 5 assignment, though your Lab 5 data will also include an id variable. Lab 5 data will be similar, if not the same, as your Lab 4 data.

First, load necessary packages:
```{r packages, message = FALSE}
needs(tidyverse, rio, here, knitr, kableExtra, misty, janitor, rstatix, psych, misty)
```

## Describe Data

Next, we will read in the data, similar to the process in your previous labs.
``` {r}
df <- misty::read.sav(here("kamden-mwavita_data/factorial_ANOVA_Bossaert_et_al.sav"), use.value.labels = TRUE) %>% 
  clean_names()
```

If you want to explore the dataset, it is good practice to view the structure of the dataset (`str(df)`)and then the top 5 rows (`head(df)`). This way, you can catch if you read in the data correctly, if there are any anomalies, etc. 

Now, we run through the basic descriptives. See Lab 4 for more information or explanations of this code, if needed. 

```{r}
table(df$sen)
table(df$country)
```

Summary table of the data with both variables:
```{r}
df %>% 
  group_by(sen, country) %>% 
  summarise(n = n()) %>%
  spread(country, n) 
```

Overall descriptive statistics.
```{r}
describe(df)
```
Make sure to interpret overall distribution stats here.

By level:
```{r}
#Output for each level - sen
describeBy(x = df$soc_part, group = df$sen, 
             mat = TRUE, data = df)

#Output for each level - country
describeBy(x = df$soc_part, group = df$country, 
             mat = TRUE, data = df)
```
Using the code above, report the mean/sd and describe the distributions for the levels of each variable.

Additionally, we can pull in what we learned last week about marginal means as another way to summarize mean values. Here, let's compare the means with each other:

```{r}
means <- df %>% 
  group_by(sen, country) %>% #the two variables
  summarise(mean_soc_part = mean(soc_part)) #mean of dependent variable

means

# Graphical representation of means
means %>% 
  ggplot(aes(country, mean_soc_part, group = sen)) +
  geom_point() +
  geom_line(aes(color = sen))
```

You can play around with the aesthetics of the graph above. There's not only one way to represent the means, just do whatever makes sense for you.

### Boxplots

```{r}
soc_boxplot <- df %>% 
  ggplot(aes(x = country, y = soc_part, fill = sen)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position="bottom")

soc_boxplot
```

## Checking assumptions

Homogeneity of variance:
```{r}
car::leveneTest(soc_part ~ country*sen, data = df, center = "mean")
```

Make sure to interpret if all assumptions are met. 


### A priori contrasts

We can do two-factor contrasts with the emmeans package.

First, we can try using *by* grouping.
```{r}
needs(emmeans)
model <- lm(soc_part ~ sen + country + sen:country, df) #remember sen, country, and sen:country interactions in the model
means_1 <- emmeans(model, specs = "sen", by = "country") #use the model to calculate means
means_1

pairs(means_1) #now display those means in pairs
```

This output shows us all the estimates by country, but it doesn't quite give us all the comparisons we need altogether. Use the code below instead.

Calculate all possible comparisons and not "nesting" one factor in another:
```{r}
means_2 <- emmeans(model, ~sen*country) #use same model but use different specifications
means_2

pairs(means_2) #display means in pairs, unnested
```

Now we have more than we need, so just pull out the values we can actually interpret (similar to what you did in Lab 4).

### Two-way ANOVA

```{r}
m1 <- anova_test(df, formula = soc_part ~ sen + country + sen:country, 
                 detailed = TRUE, type = 3, effect.size = "pes")
m1
```

### Factorial ANOVA

Let's establish some hypotheses to model the kind of analysis you'll do for Lab 5. 

* Hypothesis 1 - To what extent does the difference between typically developing and other developing types (special education, behavioral needs, other special education needs) depend on country (Belgium, Netherlands, Norway)?
* Hypothesis 2 -

This is a two-way factorial ANOVA. The code below is just a model.
```{r}
means_2

contrasts <- list(
  hyp1 = c(1, 0, -1, -1, 0, 1), #hypothetical contrasts: fill in appropriate values
  hyp2 = c(1, -.5, -.5, -1, -.5, -.5) #hypothetical contrasts: fill in
)

contrast(means_2, contrasts, adjust = "holm")
```
There is a xx word difference in vocabulary learning between physical science and humanities lectures, depending on lecture method. Additionally, there is a xx word difference when comparing traditional and computer-assisted lecture methods for physical science as compared to social science. 

### Effect Sizes

**why omega vs. pes**

Calculate effect sizes with partial eta squared.
```{r}
# presume -1.33 is the t.ratio from hyp1
# presume -2.05 is the t.ratio from hyp2

## actually prefer omega here, but use pes for the purposes of this lab

effectsize::t_to_eta2(
  t = c(-1.33, -2.05),  #hypothetical, fill in your own
  df_error = 28  #hypothetical, fill in your own
)
```
The partial eta squared of xx indicates that lecture method explains about 3% of the difference in vocabulary learning between physical science and humanities lectures. 

The partial eta squared of xx indicates that lecture method explains about 19% of the difference in vocabulary learning between physical science and social science conditions.


#### For One-way ANOVA
Going through the process above for one-way ANOVA, exploring interactions.

```{r}
m1 <- lm(vocab ~ inst, vocab5_data)
means1 <- emmeans(m1, ~inst)
means1

#Specify contrasts

contrasts1 <- list(
  hyp1_1 = c(1, -1, 0), #hypothetical contrasts, fill in your own
  hyp2_1 = c(1, 0, -1)  #hypothetical contrasts, fill in your own
  )
contrast(means1, contrasts1)

effectsize::t_to_eta2(
  t = c(3.75),  #hypothetical, fill in your own
  df_error = 12  #hypothetical, fill in your own
)
```
