---
title: "Lab 5 Key"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Study Background

In an experiment, fifth grade students are individually and randomly assigned to be introduced to a set of vocabulary words by watching a lecture in one of three subject areas: physical science, social science, and history.  The lecture was delivered in one of two methods: traditional lecture or computer-augmented lecture.  Six children experienced each combination of conditions.  A vocabulary test with 60 words follows each lecture. Our research question was: “To what extent do lecture topic and method influence the vocabulary learning of fifth grade students?”

But we now have two post-hoc interaction research questions to test: 

“To what extent does the difference between physical science and humanities lectures on vocabulary learning depend on lecture method?”
“To what extent does the difference between traditional and computer-assisted lecture methods differ for physical science as compared to social science?”

```{r}
needs(tidyverse, rio, here, knitr, kableExtra, misty, janitor, rstatix, psych, misty)

vocab5_data <- misty::read.sav(here("assignments_data", "Lab5_Vocab.sav"), use.value.labels = TRUE)
```

# 2. Examining Data and ANOVA Assumptions

This is a between-subjects analysis, as students were randomly assigned to only one type of instruction which, along with the balanced sample among groups, supports the _assumption of independence of observations_. The _assumption of normality_ seems tenable after visually inspecting the distribution and descriptive statistics for _Vocabulary Scores_. Finally, the _assumption of homogeneity of variance_ was not met as Levene's Test for Homogeneity of Variance was not significant (_F_(5, 30)= 2.06, _p_ = .099), indicating that the error (within-group) variance around the mean is equal across groups.

See the following code for assessing the assumptions.

```{r}
describe(vocab5_data)
describeBy(x = vocab5_data$vocab, group = vocab5_data$inst, 
             mat = TRUE, data = vocab5_data)
```

```{r}
means <- vocab5_data %>% 
  group_by(inst, meth) %>% 
  summarise(mean_vocab = mean(vocab))

means

means %>% 
  ggplot(aes(inst, mean_vocab, group = meth)) +
  geom_point() +
  geom_line(aes(color = meth))
```

```{r}
car::leveneTest(vocab ~ inst*meth, data = vocab5_data, center = "mean")
```


# 3. Two-way ANOVA with Interactions Contrasts

## 3.1 Marginal Means and Pairwise Comparisons
As this lab is a follow-up (post-hoc) analysis for Lab 4, we will focus on the post-hoc pairwise comparisons and probing the interaction contrasts in this section of the lab. See the following code for how to examine the marginal means and pairwise comparisons.
```{r}
needs(emmeans)

model <- lm(vocab ~ meth + inst + meth:inst, vocab5_data)

means_1 <- emmeans(model, specs = "inst", by = "meth")
means_1

pairs(means_1)
```
This is okay, but it doesn't quite give us all the comparisons we need. Use the code below instead.

Calculate all possible comparisons and not "nesting" one factor in another:
```{r}
means_2 <- emmeans(model, ~inst*meth)
means_2

pairs(means_2)
```

## 3.2 Custom Interaction Contrasts
The following code is for defining the custom interaction contrasts needed for answering the research questions for Lab 5.
```{r}
means_2

contrasts <- list(
  hyp1 = c(1, -.5, -.5, -1, .5, .5), #Hypothesis 1: Physical science vs/ humanities (social science and history) depending on lecture method
  hyp2 = c(1, -1, 0, -1, 1, 0) #Hypothesis 2: Traditional vs. computer-assisted lecture methods for physical science vs. social science
)

contrast(means_2, contrasts, adjust = "holm")
## 5.5 is difference between the differences (see graph above to help interpret)
```

### 3.2.a Effect Sizes
While the code below includes functions for $\eta_p^{2}$ and $\omega^{2}$, we will only report on $\omega^{2}$ (see textbook for justification).
```{r}
# Calculating partial eta squared for our two contrasts. The output includes estimates for both without needing to run two functions.
effectsize::t_to_eta2(
  t = c(-1.043, -2.628), # including t-values for both contrasts 
  df_error = 30
)

needs(MOTE)
# Calculating omega squared for our two contrasts.
## First contrast
### Transform t-ratio into F-ratio by squaring it: t^2 = F
(-1.043)^2
# Fvalue = 1.087849
# Put in values accordingly:
omega.F(dfm = 1, dfe = 30,
      Fvalue = 1.087849, n = 36, a = .05)

## Second contrast
### Transform t-ratio into F-ratio by squaring it: t^2 = F
(-2.628)^2
# Fvalue = 6.906384
# Put in values accordingly:
omega.F(dfm = 1, dfe = 30,
      Fvalue = 6.906384, n = 36, a = .05)
```

## 3.3 Interpreting Interaction Contrast Results
_**Research Question 1.** To what extent does the difference between physical science and humanities (social science and history) lectures on vocabulary learning depend on lecture method?_

Our findings indicate that students in the physical science and history lecture groups, did not have significantly different vocabulary test scores depending on whether the lecture was delivered using traditional or computer-augmented methods (_t_ = -1.04, _p_ = .305, $\omega^{2}$ = 0.002).

_**Research Question 2.** To what extent does the difference between traditional and computer-assisted lecture methods differ for physical science as compared to social science?_

We found significant differences between mean vocabulary scores among students in the two lecture groups (_t_ = -2.628, _p_ = .013, $\omega^{2}$ = 0.141), accounting for approximately 14.1% of variance in the model. This difference remained statistically significant after correcting for multiple comparisons using a post-hoc Bonferroni adjustment (i.e., $\alpha$ = .05/3 tests = .0167). In other words, students had similar mean vocabulary test scores within the computer-augmented lecture group among the physical science lecture (_M_ = 46 words) and social science lecture (_M_ = 38 words), but when receiving the traditional lecture, students in the physical science lecture score markedly higher (_M_ = 34 words) in comparison to their peers in the social science lecture group (_M_ = 12 words).